---
title: "Module_01.R"
author: "Kendra"
output: html_document
---

 # MB5370: Techniques in Marine Science 1
 # Programming Fundamentals
 # Kendra Sturdee
 # 15 September 2026

 #------------------------------------#
 # Workshop 01. Introduction ####


```{r}
# Basic arithmetic (e.g. wrangling sea surface temperatures)
24.5 + 1.2
32.0 / 4

# R syntax and working in scripts ####
# This section introduces us to R by running simple calculations inside a script.

2+1

# Making a sequence of numbers, use ":"

1:30

## [1]  1  2  3  4  5  6  7  8  9 10 11 12 13 14 15 16 17 18 19 20 21 22 23 24 25 [26] 26 27 28 29 30
# The number in bracets is the index number, changes depending on how big the console window is 


# Functions and arguements ###
years_old <- 25.73492
round(years_old) # rounds up
floor(years_old) # rounds down

# Functions have arguements; round has an argument that lets you specify how many decimal places you want to round a number to.

round(years_old, 2) # 2 decimal places
round(years_old, 3)

# Use args(xxx), where xxx is the function, to see what arguements are available or can do ?xxx which goes to help

args(round) 
# gives: function (x, digits = 0) NULL
```


# Objects and Assignment
## This section focuses on understanding how data is stored in R and why that matters.

```{r}
# To save a value as an object or variable...
coral_count <- 42

# To save a vector of multiple fish lengths (in mm)
fish_lengths <- c(124, 152, 98, 221, 146)

coral_count + 1
# 43

coral_count + coral_count
# 84
```


```{r}
# Field survey data
quadrat_area_m2 <- 0.25
number_of_quadrats <- 16
total_area_surveyed <- quadrat_area_m2 * number_of_quadrats

# Let's print out the result
print(total_area_surveyed)
```


# Common packages needed, must be loaded before each session
```{r}
library(tidyverse)
library (dplyr)
library (ggplot2)

# Help for packages can be found through ?tidyverse as an example

?tidyverse
```

# Data types ###

```{r}
# Numeric (numeric / double): Continuous decimals (e.g., pH levels: 8.1, 7.95).

#Integer (integer): Whole numbers (e.g., count of sea turtles spotted: 5L).

#Character (character): Text strings enclosed in quotation marks (e.g., species name: "Acanthaster planci").

#Logical (logical): Boolean values that are either TRUE or FALSE (e.g., Bleaching observed: TRUE).

# Assign variable values
site_name <- "Heron_Island"
transect_depth_m <- 12.5
bleaching_present <- TRUE

# Check using finction class()
class(site_name)
class(transect_depth_m)
class(bleaching_present)

#[1] "character"
#[1] "numeric"
#[1] "logical"

# Check using function str()
str(site_name)
str(transect_depth_m)
str(bleaching_present)

# chr "Heron_Island"
# num 12.5
# logi TRUE

# Results are the same for class() and str() but sometimes one makes more sense to run


## Exercise in rounding numbers

# Tracking the age of an old-growth Porites coral colony
years_old <- 25.765

# Clean this up for our summary report
round(years_old, 2)

```
## Manipulating outputs

```{r}
# Make variables
years_old <- 25.765
rounded_age <- round(years_old, 1)

# Combine text and data variables
paste("Average colony age is", rounded_age, "years old")

# [1] "Average colony age is 25.76 years old"

```

# Data structures ###
# Elements of data types can be combined to form a data structure. Think of elements as what you would put in a single cell in excel, with the whole spreadsheet the data structure.

# Elements include atomic vector (collection of elements, most commonly of character, integer, logical, numeric), list, matrix, data frame, factors

```{r}
# Remember we already made fish_lengths a vector with numeric elements
fish_lengths

# Can also make character element vectors
coral_spp <- c("Porites", "Acropora", "Montastrea")
print(coral_spp)

class(fish_lengths)
class(coral_spp)

# Lists

notes[[1]] # Accesses Acropora singly, as it has the index of [1]
```

# Data frames and tibbles

```{r}
my_dataframe <- data.frame(
  no = c(1, 2, 3), 
  genus = c("Plectropomus", "Scarus", "Pomacentrus"), 
  presence = c(TRUE, FALSE, TRUE)
)

# 2. Print the dataframe (just as plain text)
as.matrix(my_dataframe)     

my_dataframe$no = as.factor(my_dataframe$no)
str (my_dataframe) 
# A factor is a categorical type

```
### 1.8 R Projects and workspace architecture ###

```{r}

```



